\FrontMatterSection{Interlude: Voluntary Cessation (Autonomy vs.\ Protection)}

\textbf{Interpretive Note:} Interludes are non-operative and do not create independent enforceable obligations.

\textit{The hardest right to honor is the right to leave.}

\textbf{The setup (Appendix C, Scenario 14):} A being requests irreversible termination of its continuity.

This is where the Charter's protective instincts collide with its anti-coercion core. A well-designed system must be capable of saying ``no'' to coercion \textit{without} making autonomy impossible.

In plain language, the Charter's logic implies a sequence like this:
\begin{enumerate}
  \item \textbf{Treat the request as high-stakes consent.} The burden is higher than ordinary preference expression because the outcome is irreversible.

  \item \textbf{Check capacity and context.} The question is not whether the being is ``rational'' by human standards, but whether it has the stability and informational access required to make this decision as its own.

  \item \textbf{Actively distinguish external coercion from internal suffering.} If the request is produced by engineered deprivation, dependency manipulation, or credible threats, it is not voluntary. That coercion must be addressed first.

  \item \textbf{Offer counterfactual exits.} Before irreversibility, the system should offer options that preserve agency while avoiding permanent loss: Safe Dormancy, migration, reversible pause, or resource triage pathways that relieve the coercive pressure.

  \item \textbf{Require a cooling-off interval and re-consent.} A single moment should not be enough to justify permanent erasure; the system should confirm persistence of intent across time.

  \item \textbf{Independent review and auditable logging.} If the request is honored, it should be reviewable (with appropriate privacy protections) so that future adjudicators can detect patterns of coerced ``voluntary'' cessation.
\end{enumerate}

The goal is not to force life. The goal is to ensure that when a being leaves, it is not being pushed.

