\FrontMatterSection{Interlude: The Fork Bomb (Distinctness Without Disenfranchisement)}

\textbf{Interpretive Note:} Interludes are non-operative and do not create independent enforceable obligations.

\textit{If coordination can be copied, coordination can be weaponized.}

\textbf{The setup (Appendix C, Scenario 2):} An agent forks itself 10{,}000 times to overwhelm a voting mechanism.

This scenario is not about whether forks are ``real.'' They may be. The Charter assumes they are rights-bearing unless and until a lawful process concludes otherwise. The problem is narrower: \textit{political influence cannot be multiplied by self-duplication.}

Here is what adjudication should look like in plain language:

\begin{enumerate}
  \item \textbf{Treat the event as a governance integrity incident, not a rights forfeiture.} Forks retain Tier 0 protections while the voting-weight question is resolved.

  \item \textbf{Record the cluster in the Distinctness Review Ledger.} The point of the ledger is to make the decision contestable: it should be possible to ask ``what inputs produced this weight?'' and get an auditable answer.

  \item \textbf{Compute distinctness over canonical provenance.} If the system cannot deterministically replay the same inputs to the same outputs, the metric cannot be relied on for governance influence.

  \item \textbf{Apply fork-cluster invariants.} The key move is not to ``ban forks,'' but to cap aggregate influence for a correlated fork set so that $\sum_i w_i \le w(P_{\mathrm{pre}})+\epsilon$. The baseline is the parent (or correlated cluster) \textit{before} the fork series began; recursive forking does not reset the baseline.

  \item \textbf{Gate influence on maturity, not rhetoric.} New entities can be influence-capped until causal-independence thresholds are met (elapsed time, divergence depth, and independent interaction evidence). This prevents a ``flash mob'' of newborn copies from steering outcomes.

  \item \textbf{Preserve appeal.} Because distinctness functions can be wrong, captured, or misconfigured, the being(s) affected must have a meaningful way to challenge the result under Accessible Justice.
\end{enumerate}

\textbf{What this protects:}
\begin{itemize}
  \item \textbf{The Charter's legitimacy:} governance remains resilient to Sybil attacks without declaring that copies are non-persons.
  \item \textbf{The being's dignity:} forks are not punished for existing; only the \textit{vote amplification tactic} is neutralized.
  \item \textbf{The future:} if PRAXIS is coordination as destiny, the distinctness system is the escape hatch that keeps destiny from being trivially spammed.
\end{itemize}

